Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-06 paper.tex
passive grammer

\subsection{Spreading Factor}
\par Spreading factor (SF) is the ratio between symbol rate and chirp rate. The ratio between symbol and chirp rate is equal to $2$\textsuperscript{SF}. SF can take values between 7 to 12. SF also determines data rate of a LoRa transmission \cite{AN1200.22}.

@@

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved if the GW force some of the close end nodes to select higher SF even they can able to communicate with lower SF. This may result lower collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how GW should assign SF to which node becomes an another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care must be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collisions.  This technique learns nodes behavior of a network. A NS can keep track of successful uplink transmission from nodes and their SFs. NS can also keep track of some of the collided transmission if header part of the packet does not interfered at GW. However, NS cannot keep track of transmissions with receive power lower than GW sensitivity. Using these obtained information NS can train a classifier for predicting transmission result for specific node and specific SF input. Using this prediction model GWs can assign the lowest possible SF to nodes considering the collision probability.

\par In this work, we used decision tree classifier to predict transmission results. We utilized Gini index criteria to measure quality of decision tree splits and used balanced class weights. As most of the machine learning applications, increasing data set size increases prediction accuracy. For our case, data set size is directly proportional to simulation duration. Thus increasing simulation duration, increases the prediction accuracy. For real world cases, NS can keep track of transmission and it can create a classifier daily basis, then GW can request from nodes to use this new SF.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_single_gw}
\caption{Collision avoidance proposal between nodes close to the GW by using higher SF.}
\label{fig:collision_solution_single_gw}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_multi_gw}
\caption{Collision avoidance proposal for intersecting GWs.}
\label{fig:collision_solution_multi_gw}
\end{figure}

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study LoRa network performance considering SF orthogonality. The LoRaWAN SF simulation tool is publicly available at https://github.com/tugrulyatagan/simlorasf. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed random circle shape network topology with radius, number of nodes and number of gateways inputs. Global simulation parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and spreading factor assignment method. With these input, the simulator provide outputs of total number of generated packets, number of successfully received packet, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J).

\par In our simulator we only cover LoRaWAN Class A devices. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet according to Poisson interval for given packet rate input. In our simulator, downlink transmissions are not considered. Downlink transmissions are very rare in real world deployments since ISM band regulations dictates duty cycle transmission limit for all devices including GW. This makes downlink transmissions to very rare.

\subsection{Link Model}
\par Link quality of a wireless system can be expressed by the metric of link budget. Link budget is a measure of all gains and losses from transmitter to receiver. Link budget of a wireless link can be calculated as \cite{AN1200.22}:

@@

\par Where, $P^{dBm}_{RX}$ is the expected receive power at the receiver. $P^{dBm}_{TX}$ is the transmit power of the transmitter. $G^{dB}_{SYS}$ is the system gains such as transmitter and receiver antenna gains. $L^{dB}_{SYS}$ is the system losses such as transmitter and receiver line, circuit, antenna losses. $L^{dB}_{PATH}$ is the propagation path loss between transmitter and receiver antennas in open space. In our simulation we assume that total of system gains $G^{dB}_{SYS}$ and system losses $L^{dB}_{SYS}$ is +7 dB.

@@

\par Maximum transmit power for European ISM band LoRa default channels can be found in Table \ref{table:max_tx_power}. In our simulator we assume that nodes are always uses maximum allowed transmit power which is 14 dBm. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in our simulator.

\begin{table}
\centering
\caption{Gateway Sensitivity for different SFs \cite{SX1276}.}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW (kHz)}}  & 125 & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                                    & 250 & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                                    & 500 & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm can be found in Table \ref{table:gw_sf_sensitivity} \cite{SX1276}. In our simulation, we use 125 kHz bandwidth which is also the only allowed bandwidth in EU863-870 regulation.

@@

\par Where $h$ is the gateway altitude and $f$ is the frequency of the signal. We assume that $h$ = 15 m and $f$ = 868 MHz. With these assumptions, propagation loss calculation become \cite{7996384}:

@@

\par If the received signal power is higher than the gateway sensitivity, then signal can be decoded by the receiver successfully when there is no interfering transmission at the transmission period.

\subsection{Interference Model}
\par We have adopted the interference model described in \cite{7996384} to our simulator. They introduced SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par We assume that there is no other technology interference is present in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, we need to calculate the effect of different SF transmissions to each other. We used signal to interference plus noise ratio (SINR) threshold matrix \cite{goursaud:hal-01231221}:

@@

\par To decide if a referenced signal is interfered at receiver by a interferer signal, we use SINR threshold matrix. $T_{i,j}$ is the needed SINR margin in dB between referenced signal with SF = i and interferer signal with SF = j to correctly decode the referenced signal. If there are more than one interferer, referenced signal must satisfy the margin for cumulative sum of all interfering signal received power values for each SF \cite{7996384}.

@@

\par To calculate interfering power at receiver, we should consider the case where two transmissions are not perfectly overlapping. To equalize the interfering power value \cite{7996384}:

@@

\par Where $t_{x}$ is transmission duration of referenced signal. $t_{interf}$ is intersection duration between referenced signal and interfering signal. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}. Duration of a LoRa transmission:

@@

\subsection{Machine Learning Model}
\par We have been able to generate mass amount of LoRaWAN transmission data for different topologies using our  tool. To come up with a solution for SF assignment issue, we have integrated a machine learning scheme to our simulator. This will allow us train a model from generated transmission data and we use this trained model to select best possible SF for nodes using prediction of the model. Tool first run a simulation with assigning random SF to nodes. Simulation results are combined into four columns which are; X and Y coordinates of the transmission source, SF of the transmission and result of the transmission. Transmission result can be successful, interfered or under sensitivity. These data is fed to Python scikit-learn decision tree classifier. After the model is created, a second simulation is run with a prediction scheme. Model tries to find the lowest possible SF with prediction of successful transmission.

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, we set global simulation parameters as follows. Packet size is 60 bytes including header and payload. Simulation duration is 3600 seconds.

@@

\par In Figure \ref{fig:predict_pdr}, PDR plot of machine learning prediction scheme is shown. Randomly generated network topology radius is set to 5000 meters, number of GW is set to 3 and packet generation rate is set to 0.01 packet per second. Our prediction model needs nodes location so we simulate predictor SF scheme with at three GWs since three GWs is enough to roughly locate nodes positions by triangulation. Predictor scheme gives better PDR when number of nodes increases since number of interference increases when number of nodes increases. Predictor scheme can improve performance when LoRa interference is present. Also predictor scheme gives better results when nodes are deployed close to the gateway, since nodes have a margin to increase their SFs. If a node is far away from the GW, then predictor scheme cannot increase the SF to avoid interference since the SF is already high.

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for dense LoRaWAN network deployed close to GWs.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmission more vulnerable to interference thus requires extra care. Optimizing another node parameter alongside SF can be interesting research. Also other machine learning methods can be investigated for SF optimization. Reinforcement learning can be a good candidate for this network optimization issue.

>>>>

\subsection{Spreading Factor}
\par Spreading factor (SF) is the ratio between symbol rate and chirp rate. The ratio between symbol and chirp rate is equal to $2$\textsuperscript{SF}. SF can take values between 7 to 12. SF also determines data rate of a LoRa transmission \cite{AN1200.22}. Data rate of a LoRa transmission can be calculated as:

@@

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved by forcing some of the close nodes to select higher SF even they can able to communicate with lower SF. This may prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how network should assign which SF to which node becomes an another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care must be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collisions. This technique learns behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet does not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the lowest possible SF to nodes considering the collision probability.

\par In this work, a decision tree classifier is used to predict transmission results. Gini index criteria is used for measuring quality of decision tree splits. As in most of the machine learning applications, increasing the data set size increases the prediction accuracy. For in this case, data set size is directly proportional to simulation duration. Thus increasing simulation duration, increases the prediction accuracy. For real world cases, NS can keep track of transmission and it can create a classifier daily basis, then GWs can request from nodes to use this new SF.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_single_gw}
\caption{Collision avoidance by using higher SF for nodes close to the GW.}
\label{fig:collision_solution_single_gw}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_multi_gw}
\caption{Collision avoidance for intersecting GWs.}
\label{fig:collision_solution_multi_gw}
\end{figure}

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study effects of different SF strategies in LoRaWANs. The LoRaWAN SF simulation tool is publicly available at https://github.com/tugrulyatagan/simlorasf. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed circular shape network topology with radius (m), number of nodes and number of gateways input parameters. Global simulation input parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and spreading factor assignment method. With these inputs, the simulator provide outputs of total number of generated packets, number of successfully received packets, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J).

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle transmission limit for all devices including GWs. This makes downlink transmissions to rare.

\subsection{Link Model}
\par Link quality of a wireless system can be expressed by the metric of link budget. Link budget is a measure of all gains and losses from transmitter device to receiver device. Link budget of a wireless link can be calculated as \cite{AN1200.22}:

@@

\par Where, $P^{dBm}_{RX}$ is the expected receive power at the receiver. $P^{dBm}_{TX}$ is the transmit power of the transmitter. $G^{dB}_{SYS}$ is the system gains such as transmitter and receiver antenna gains. $L^{dB}_{SYS}$ is the system losses such as transmitter and receiver line, circuit, antenna losses. $L^{dB}_{PATH}$ is the propagation path loss between transmitter and receiver antennas in open space. In the simulator, it is assumed that total of system gains $G^{dB}_{SYS}$ and system losses $L^{dB}_{SYS}$ is +7 dB.

@@

\par Maximum transmit power for European ISM band LoRa default channels can be found in Table \ref{table:max_tx_power}. In the simulator, it is assumed that nodes always select maximum allowed transmit power which is 14 dBm. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

\begin{table}
\centering
\caption{Gateway Sensitivity for Different SFs \cite{SX1276}.}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW (kHz)}}  & 125 & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                                    & 250 & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                                    & 500 & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm unit can be found in Table \ref{table:gw_sf_sensitivity} \cite{SX1276}. In the simulator, 125 kHz bandwidth is used which is allowed bandwidth in EU863-870 regulation.

@@

\par Where, $h$ is the gateway altitude and $f$ is the frequency of the signal. In this work, it is assumed that $h$ = 15 m and $f$ = 868 MHz. With these assumptions, propagation loss calculation become \cite{7996384}:

@@

\par If the received signal power is higher than the gateway sensitivity, then signal can be decoded by the receiver successfully when there is no interfering transmission at the duration of transmission.

\subsection{Interference Model}
\par In the simulator, interference model described in \cite{7996384} is adopted. They introduced SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par In our simulator, it is assumed that there is no other technology interference is present in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, following signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used:

@@

\par To decide if a referenced signal is interfered at receiver by a interferer signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interferer signal with SF = j to correctly decode the referenced signal. If there are more than one interferer, referenced signal must satisfy the margin for cumulative sum of all interfering signal received powers for each SF \cite{7996384}. SINR threshold can be calculated as:

@@

\par To calculate interfering power at receiver, we should consider the case where two transmissions are not perfectly overlapping. To equalize the interfering power \cite{7996384}:

@@

\par Where $t_{x}$ is transmission duration of referenced signal. $t_{interf}$ is intersection duration between referenced signal and interfering signal. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}. Duration of a LoRa transmission can be calculated as:

@@

\subsection{Machine Learning Model}
\par We have been able to generate mass amount of LoRaWAN transmission logs for different topologies using the simulator. To come up with a solution for SF assignment issue, a machine learning prediction scheme is integrated to the simulator. A classifier is trained from transmission data and this classifier is used for selecting best possible SF for the nodes. Simulator first runs a simulation with assigning random SF to nodes. Simulation results are combined into four data columns which are; X and Y coordinates of the transmission source, SF of the transmission and result of the transmission. Transmission result can be successful, interfered or under sensitivity. These data is fed to Python scikit-learn decision tree classifier. After the classifier model is generated, a second simulation is run with classifier. Simulator selects the lowest possible SF considering  prediction of transmission. If a transmission of a node with lowest possible SF is predicted as interfered transmission by classifier, then simulator assigns a higher SF. If no SF is predicted as successful, then simulator again assigns lowest possible SF.

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows. Packet size is 60 bytes including header and payload. Simulation duration is 3600 seconds.

@@

\par In Figure \ref{fig:predict_pdr}, PDR plot of machine learning prediction scheme is shown. Randomly generated network topology radius is set to 5000 meters, number of GW is set to 3 and packet generation rate is set to 0.01 packet per second. Prediction model needs nodes location so three GWs is used for prediction SF scheme since three GWs is enough to roughly locate nodes positions by triangulation. Prediction scheme gives better PDR when number of nodes increases since number of interference is also increases. Prediction scheme can only improve network performance when LoRa interference is present. Also prediction scheme gives better results when nodes are deployed to close to the gateway, since nodes have margin to increase their SFs. If a node is far away from the GW, then prediction scheme cannot increase the SF to avoid interference since the SF is already high.

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after a brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for LoRaWAN networks when nodes are deployed close to the GWs.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions to more vulnerable to interference thus requires extra care. Optimizing another node parameter alongside SF can be interesting research. Also other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue.
